Writing Trash Bin

### 2024-03-09 paper.qmd
updated the similar work section

## Similar work
::: aside
I have a large backlog of uncertainty literature review summaries in my Mendely that need to be added here. That is very high on my todo list once the readable draft is done.
:::

Any field that experiences an explosion of new research will also invite a series of literature reviews that seek to combine and summarise that information. Given that uncertainty visualisation has been around for a few decades now, there is already a wealth of uncertainty visualisation literature reivews. Interestingly these papers seem to have identified a series of common pitfalls in uncertainty visualisation and its evaluation.

@Kinkeldey2014 summarised the literature specifically for geopsatial uncertainty visualisations and attempted to describe similarities in their characteristics. They found most experiments the methods for uncertainty visualisation evaluation to be adhoc, with no commonly agreed upon methodology or formalisation. They found that experiments took on an engineering approach, where plots were evaluated with the goal to improve a particular representation, rather than to investigate why particular representations do or do not work and create general principals. Authors would then ignore the engineering perspective they had taken on, and try to generalise beyond the specific constraints of their experiment. While they did discuss the results of some perceptual task experiments they found that even those results were somewhat conflicting and difficult to combine in any meaningful way. They also stress the importance of selecting the correct audience, as numerical and visual literacy can differ vaslty between subpopulations so the results from one populations may not reflect the results from another. The paper contains an entire paragraph where experiments that compared box-plot representations are rattled off, each experiment finding results that conflicted with its predecessor. Ultimately they suggested the visualisation typologies move away from data types, uncertainty categories, and representation types and towards "task-centered typologies" to better enable generalisations of the findings.

@Hullman2016 reviewed the uncertainty visualisation literature more broadly (not specifically for geospatial uncertainty) and offered advice on methodology and analysis of results for other uncertainty visualisation experimenters. She found similar issues in the literature and also noticed the somewhat ad-hoc methodology though the lack of justification in the method for the particular questions participants were required to answer, and the lack of generalisability of comparing plots one by one to create a ranking. She also discussed the role of heuristics in the analysis of results and suggested authors look for signs of them in responses and ask subjects to describe their strategy in their response. What was particularly interesting was people's tendancy to ignore uncertainty in a plot all togehter, and use a pattern in the signal (such as distance between means) to answer questions that require uncertainty information. She also discussed the lack of clarity in exactly what "uncertainty" authors are trying to ask about. Whether the participants are supposed to answer according to some ground truth or their subjective probability is unclear, as is whether the *uncertainty* depicted represents the uncertainty of a sampling or population distribution. Ultimately all these issues become confliated and it becomes impossible to tell if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a heuristic.

@Spiegelhalter2017 presents an literature reivew of uncertainty communication as a whole and considers uncertainty visualisation as a subfield of communication as a whole. In this review he has several key findings that are of particular interest. First, the best visualisation is highly dependent on your audience, and depending on how broad your audience is, there may not be a best visualisation. No single representation will be accessible to all members of an audience, and even if most audience members can read a plot, they may not read it in the same way. For example, people with high numeracy skills count pictographs, but people with low numeracy assess area, and seemingly unrelated elements may not equally influence all these methods of reading, for example scattering in a pictograph makes counting particularly more difficult. For these reasons, it is better to understand your audience and iterate towards a final design, rather than only consider one option. Second, the best visualisation for one task is unlikely to be the best visualisation for another task. Graphics are good for "gists" and different graphics are good for different "gists", e.g. one plot may we good at identifying trend, while another might be good for comparison. Therefore, if we change the goal of our visualisation, and therefore the metric we are trying to evaluate, we implicitly change the best visualisation as no visualisation will both cover and summarise enough information to be the best at answering every question. Finally, the visualisation advice he gives is more specifically about general visualisation practice rather than advice specifically for uncertainty visualisation. Common sentiments such as avoiding chart junk, considering a table in place of a visualsiaiton, use of narrative labels, managing emotion, less is more approach, etc, are the main concepts expressed. The only uncertainty specific advice given is that the graphic should allow "part-to-whole" comparison on an appropriate scale, which seems to be an extension of the framing concept discussed when speaking about uncertainty communication more broadly.

@Griethe2006 provided a summary of the uncertainty visualisation literature and did not take any particular angle in the analysis. She makes the point that there is no universally accepted definition of uncertainty and no unicersally accepted way to display it. She defined uncertainty as "the degree to which the lack of knowledge about the amount of error is responsible for hesitancy in accepting results and observations without caution". She then explains that there are two ways to visualise uncertianty, implicitly and explicitly. When a visualisaion is implicit, the uncertainty is considered an uninteresting feature of the data, and instead of mapping some quantification of uncertainty to a visual feature, data is filtered out if it does not meet a specified "certainty" threshold. These statements clearly points towards a signal/noise dichotomy with normal visualisation and uncertainty visualisation. Additionally she notes that the motivation of the analysis is important, by stating "one has to answer the questions which kinds of uncertainty are relevant from the data’s perspective, which kinds are helpful from the perspective of the visualization steps and which kinds are practicably extractable from the perspective of the available methods?", and "The decision of which of these techniques to choose strongly depends on the intended goal" when commenting on which uncertainty visualisation techniques should be used. She clearly sees uncertainty as additional information or data, and therefore sees the uncertainty visualisation problem as a specific case of the high dimensional visualisation issues. She mentions that there are detailled classifications of uncertainty but the literature and subsequent techniques are not ballanced, and instead concentrate on scalar values. In facing this lack of ballance she creates a new classification system for uncertainty visualisation. She commented that uncertainty appears at every step in the analyis pipeline, and commented on some the mathematical definitions that seek to quantify uncertianty, but did not explain how these mathematical definitions can be combined across levels. Unfortunately this quantification is needed since she mentioned that in order to visualise uncertainty, it needs to be measurable.

These literature reviews highlight a common theme, the field of uncertainty visualisation is near impossible to navigate. There is so much noise in the experimental results that general patterns are difficult to identify and extract for a common purpose. This noise is attributed to several things, the results dependence on tasks, the difficulty in understanding the specifics of uncertainty and the conflation of signal and noise from many steps in the experimental design. Additionally these reviews often try to look at uncertainty visualisation from a specific perspective (e.g. geospatial or bayesian modelling) and find that even within these sub-fields the data is too noise to come to any meaningfull conclusions.

In contrast to these literature reviews, we do not have a specific domain or tool analysis we intend to perform. Rather, the point of this literature review is to understand the *purpose* of visualising uncertainty, and understand if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for improvement. Additionally, rather than discussing "uncertainty visualisation" as its own field, we view it as a subfield of both uncertainty and visualisation, and understand if this change in conceptualisation can alleviate or help us to find some signal in the noise of the field. Ultimately, this paper should serve as a guide for future uncertaitny visualisation authors to prevent their work from becoming noise in what is already an incredibly noisy field.

>>>>

## Similar work
::: aside
I have a large backlog of uncertainty literature review summaries in my Mendely that need to be added here. There are a lot of them.
:::

Any field that experiences an explosion of new research will also invite a series of literature reviews that seek to combine and summaries that information. Given that uncertainty visualisation has been around for a few decades now, there is already a wealth of uncertainty visualisation literature reviews.

Interestingly, these reviews often do not offer overarching rules for tried and tested uncertainty visualisation, but rather comment on the *difficulties* faced when trying combine the papers from this field. @Kinkeldey2014 found most experiments the methods for uncertainty visualisation evaluation to be adhoc, with no commonly agreed upon methodology or formalisation and no greater goal of describing general principals. @Hullman2016 commented on the difficulty in taking overarching themes from uncertainty visualisation, as several conflated issues make it unclear if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a heuristic. @Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal "best" plot for all people and circumstances. @Griethe2006 was unable to find common themes, but instead listed the findings and opinions of each uncertainty visualisation paper.

There are a handful of reasons that are used to explain why it is so difficult to generalise the findings in the uncertainty visualisation literature. Some papers suggested or agree with the idea that visualisation typologies should move away from data types, uncertainty categories, and representation types and towards "task-centered typologies" as this will help generalise results [@Kinkeldey2014; @Hullman2016]. Some authors may not attribute the task to the *reason* the literature cannot be easily summarised, but do mention that the choice of best visualisation is highly dependent on the specific goal of that visualisation [@Griethe2006; @Spiegelhalter2017]. Others comment that visualisations are highly dependent on the audience and there is no such thing as a "best" visualisation that will be accessible to all members [@Kinkeldey2014; @Spiegelhalter2017]. These difficulties in noisy uncertainty visualisation findings arises reguardless of the direction the analysis was approached, be it looking at geospatial data [@Kinkeldey2014], or uncertainty visualisation as a whole [@Hullman2016; @Spiegelhalter2017; @Griethe2006].

This review believe all these sources of noise have a single root cause, a lack of a cohesive and encompassing definition of uncertainty. The lack of an econompasing definition is touched on by every uncertainty visualisation directly [@Spiegelhalter2017; @Griethe2006] or indirectly by describing a miriad of ways it can be considered in the literature [@Kinkeldey2014; @Hullman2016], although it is never commented on as a source of the noise in the field.

In understanding the definition problem, we will understand the *purpose* of visualising uncertainty, and understand if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for improvement. In understanding this purpose, we will also consider whether or not "uncertainty visualisation" is a valid sub-field of visualisation itself, or if that distinction has only arisen as a by product of the confused definition of uncertainty. Ultimately, this paper should serve as a guide for future uncertaitny visualisation authors to prevent their work from becoming noise in what is already an incredibly noisy field.

### 2024-03-09 writingdumpster.qmd
updated the similar work section

>>>>

- @Griethe2006 clearly sees uncertainty as additional information or data, and therefore sees the uncertainty visualisation problem as a specific case of the high dimensional visualisation issues. 
- @Griethe2006 mentions that there are detailled classifications of uncertainty but the literature and subsequent techniques are not ballanced, and instead concentrate on scalar values. In facing this lack of ballance she creates a new classification system for uncertainty visualisation. 
- @Griethe2006 commented that uncertainty appears at every step in the analyis pipeline, and commented on some the mathematical definitions that seek to quantify uncertianty, but did not explain how these mathematical definitions can be combined across levels. Unfortunately this quantification is needed since she mentioned that in order to visualise uncertainty, it needs to be measurable.
- @Griethe2006 defined uncertainty as "the degree to which the lack of knowledge about the amount of error is responsible for hesitancy in accepting results and observations without caution". 
- @Griethe2006 found that there there are two ways to visualise uncertianty, implicitly and explicitly. When a visualisaion is implicit, the uncertainty is considered an uninteresting feature of the data, and instead of mapping some quantification of uncertainty to a visual feature, data is filtered out if it does not meet a specified "certainty" threshold. These statements clearly points towards a signal/noise dichotomy with normal visualisation and uncertainty visualisation.
- @Spiegelhalter2017 even if most audience members can read a plot, they may not read it in the same way. People with high numeracy skills count pictographs, but people with low numeracy assess area, and seemingly unrelated elements may not equally influence all these methods of reading, for example scattering in a pictograph makes counting particularly more difficult. *also add that perceptual task audience thing*
- @Spiegelhalter2017 it is better to understand your audience and iterate towards a final design, rather than only consider one option.
- @Spiegelhalter2017 common sentiments such as avoiding chart junk, considering a table in place of a visualsiaiton, use of narrative labels, managing emotion, less is more approach, etc, are the main concepts expressed. - @Spiegelhalter2017 suggests that the graphic should allow "part-to-whole" comparison on an appropriate scale, which seems to be an extension of the framing concept discussed when speaking about uncertainty communication more broadly.
- @Hullman2016 commented on the lack of justification of particular questions used in experiments and the lack of generalisability in using plot ranking as a response technique.
- @Hullman2016 discussed the lack of clarity in exactly what "uncertainty" authors are trying to ask about. Whether the participants are supposed to answer according to some ground truth or their subjective probability is unclear, as is whether the *uncertainty* depicted represents the uncertainty of a sampling or population distribution. 
- @Hullman2016 discussed the role of heuristics in the analysis of results and suggested authors look for signs of them in responses and ask subjects to describe their strategy in their response. 
- @Hullman2016 What was particularly interesting was people's tendancy to ignore uncertainty in a plot all together, and use a pattern in the signal (such as distance between means) to answer questions that require uncertainty information. 
- @Kinkeldey2014 stress the importance of selecting the correct audience, as numerical and visual literacy can differ vaslty between subpopulations so the results from one populations may not reflect the results from another. 
